---
title: "ggtrace"
output:
  md_document:
    toc: true
    toc_depth: 2
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
ragg_png = function(...) {
  ragg::agg_png(..., res = 300, units = "in")
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "ragg_png",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# **{ggtrace}**

#### **Programmatically explore, debug, and manipulate ggplot internals**

<!-- badges: start -->
[![](https://img.shields.io/badge/devel%20version-0.4.8.9000-gogreen.svg)](https://github.com/yjunechoe/ggtrace)
<!-- badges: end -->

### **Installation**

You can install the development version from [GitHub](https://github.com/yjunechoe/ggtrace/) with:

``` r
# install.packages("remotes")
remotes::install_github("yjunechoe/ggtrace")
```

```{r setup}
library(ggtrace) # v0.4.8.9000
```

More on the 📦 package website: [https://yjunechoe.github.io/ggtrace](https://yjunechoe.github.io/ggtrace)

## **Extending `base::trace()` with `ggtrace()`**

The low-level function `ggtrace()` is designed for interacting with functions and ggproto methods in the `{ggplot2}` ecosystem, from the "outside". 

Formally put, `ggtrace()` allows the user to inject arbitrary expressions (called **traces**) to functions and methods that are evaluated over the execution of a ggplot. When "triggered" by the evaluation of the ggplot, these traces may modify the resulting graphical output, or they may simply log their values to the "tracedump" for further inspection by the user. Check out the [FAQ vignette](https://yjunechoe.github.io/ggtrace/articles/FAQ.html) for more details.

Briefly, there are three key arguments to `ggtrace()`:
  - `method`: what function/method to trace
  - `trace_steps`: where in the method body to inject expressions
  - `trace_exprs` what expressions to inject in which step

A simple example:

```{r}
dummy_fn <- function(x = 1, y = 2) {
  z <- x + y
  x <- 10
  return(z)
}
dummy_fn()
```

Essentially, `ggtrace()` allows you to inject code into a function and safely/temporarily change its execution behavior. 

The following code injects the code `z <- z * 10` right as `dummy_fn` enters the third "step" in the body. Note that the value of `trace_exprs` must be of type "language", which we're simply calling "code" here. The idea is to give `ggtrace()` code for it to evaluate inside the function when it is ran later. Most often, typing out the actual code and wrapping it in `quote()` suffices, but for more complex injections see the [Expressions chapter of Advanced R](https://adv-r.hadley.nz/expressions.html)

```{r}
ggtrace(
  method = dummy_fn,
  trace_steps = 3,
  trace_exprs = quote(z <- z * 10)
)
```

After this `ggtrace()` call, the next time `dummy_fn` is called it is run with this injected code.

```{r}
dummy_fn()
```

Essentially, `dummy_fn` ran with this following modified code just now:

```{r}
dummy_fn_TRACED1 <- function(x = 1, y = 2) {
  z <- x + y
  z <- z * 10 #< Look here!
  x <- 10
  return(z)
}
dummy_fn_TRACED1()
```

But how do we know that's the line targeted by `trace_steps = 3` in our `ggtrace()` call?

The function `ggbody()` returns the body of a function/method as a list, which allows you to inspect what expressions in the body are evaluated at which step. `ggtrace()` injects the expression right before entering the step specified in `trace_steps`, so by setting that value to `3` we tell it to inject `z <- z * 10` right before `ggbody(dummy_fn)[[3]]` is evaluated.

```{r}
ggbody(dummy_fn)

ggbody(dummy_fn)[[3]]
```

Note that once a trace placed by `ggtrace()` is triggered, the original function/method is restored (i.e., the trace is removed). You can also check whether a function is currently being traced with `is_traced()`.

```{r}
dummy_fn()
is_traced(dummy_fn)
```

To not have a trace clean up after itself, you can set `once = FALSE` which makes a trace **persistent**, meaning that it will be there until explicitly removed with `gguntrace()`

```{r}
ggtrace(
  method = dummy_fn,
  trace_steps = 3,
  trace_exprs = quote(z <- z * 10),
  once = FALSE #< here!
)

dummy_fn()
dummy_fn()

is_traced(dummy_fn)

gguntrace(dummy_fn)
is_traced(dummy_fn)
```

Beyond these basic capabilities, `ggtrace()` offers two more features:

### **1. Shared state across trace expressions**

The values of `trace_steps` and `trace_exprs` can be of length > 1

Here, we multiply `z` by 10 right as `dummy_fn` enters Steps 3 and 4

```{r}
ggtrace(
  method = dummy_fn,
  trace_steps = c(3, 4),
  trace_exprs = quote(z <- z * 10)
)
dummy_fn()
```

The function just ran with the following modified code:

```{r}
dummy_fn_TRACED2 <- function(x = 1, y = 2) {
  z <- x + y
  z <- z * 10 #< Look here!
  x <- 10
  z <- z * 10 #< Look here!
  return(z)
}
dummy_fn_TRACED2()
```

Thus, even though we injected multiple expressions, they are evaluated in the same environment and thus are part of a single trace created by `ggtrace()`.

Also, note that `trace_exprs` was recycled here to meet the length of `trace_steps`. An explicit spellout would look like the following:

```{r, eval = FALSE}
ggtrace(
  method = dummy_fn,
  trace_steps = c(3, 4),
  trace_exprs = list(
    quote(z <- z * 10),
    quote(z <- z * 10)
  )
)
```


### **2. Logging of evaluated trace expressions**

The output of injected expressions are logged to "tracedumps", of which there are two.

First, `last_ggtrace()` stores the output of the last trace. Here, we evaluate `x` upon entering `dummy_fn` (Step 1), and right before it returns (Step 4). Note that we can target the last step with negative indices like `-1`, which count backwards from the last step.

```{r}
ggtrace(
  method = dummy_fn,
  trace_steps = c(1, -1),
  trace_exprs = quote(x),
)
dummy_fn()
```

Of course the return value has not changed, but we can inspect the output of evaluating `x` in these two places with `last_ggtrace()`:

```{r}
last_ggtrace()
```

If `trace_exprs` is a named list of expressions, the tracedump will carry those names as well

```{r}
ggtrace(
  method = dummy_fn,
  trace_steps = c(1, -1),
  trace_exprs = list(
    "x_after_begin" = quote(x),
    "x_before_end"  = quote(x)
  ),
)
dummy_fn()

last_ggtrace()
```

Second, `global_ggtrace()` stores the output of all traces. This is particularly useful in conjunction with `once = FALSE` or when you want to inspect multiple `ggtrace()`s at once. For most usecases you don't need to go this far, but should you need it the [documentation page](https://yjunechoe.github.io/ggtrace/reference/global_ggtrace.html) has more information.


## **Workflows for interacting with ggplot internals**

```{r}
library(ggplot2)
```

Admittedly, `ggtrace()` is a bit too clunky for interactive explorations of ggplot internals. To address this, we offer "workflow" functions in the form of `ggtrace_{action}_{value}()`. These are grouped into three workflows: Inspect, Capture, and Highjack.

**NOTE**: Making the most out of these workflow functions requires a hint of knowledge about ggplot internals, namely the fact that ggproto objects like [Stat](https://ggplot2.tidyverse.org/reference/ggplot2-ggproto.html#stats) and [Geom](https://ggplot2.tidyverse.org/reference/ggplot2-ggproto.html#geoms) exists, and that these [ggprotos have methods](https://ggplot2-book.org/spring1.html#methods) that step in at different parts of the ggplot build/render pipeline to modify the data. If you are completely new to these concepts, you should at least watch [Thomas Lin Pedersen](https://twitter.com/thomasp85)'s talk on [Extending your ability to extend ggplot2](https://www.rstudio.com/resources/rstudioconf-2020/extending-your-ability-to-extend-ggplot2/) before proceeding.

### **Walkthrough with `geom_smooth()`**

Say we want to learn more about how `geom_smooth()` layer works, exactly

```{r}
class(geom_smooth())

geom_smooth()
```

To do this, we're going to borrow the example from the [ggplot2 internals chapter of the ggplot book](https://ggplot2-book.org/internals.html)

```{r}
p <- ggplot(mpg, aes(displ, hwy, color = drv)) + 
  geom_point(position = "jitter") +
  geom_smooth(method = "lm", formula = y ~ x) + 
  facet_wrap(vars(year)) + 
  ggtitle("A plot for expository purposes")
p
```

Let's start with the Stat ggproto. We see that `geom_smooth()` uses the `StatSmooth` ggproto

```{r}
class(geom_smooth()$stat)
identical(StatSmooth, geom_smooth()$stat)
```

The bulk of the work by a Stat is done in the `compute_*` family of methods. We'll focus on `compute_group` here, which looks like the following:

```{r}
get_method(StatSmooth$compute_group)
```

### **Inspect**

Here we introduce our first workflow function `ggtrace_inspect_n`, which takes a ggplot as the first argument and a ggproto method as the second argument, returning the number of times the ggproto method has been called in the ggplot's evaluation:

```{r}
ggtrace_inspect_n(x = p, method = StatSmooth$compute_group)
```

As we might have guessed, `StatSmooth$compute_group` is called for each fitted line (each group) in the plot

As we can saw from `get_method()`, the `compute_group` is just a good ol' function. But what does this function actually return?

We can answer that with another workflow function `ggtrace_inspect_return()`, which shares a similar syntax:

```{r}
return_val <- ggtrace_inspect_return(x = p, method = StatSmooth$compute_group)
nrow(return_val)
head(return_val)
```

Note that `ggtrace_inspect_return()` only gave us 1 dataframe, corresponding to the return value of `StatSmooth$compute_group` the _first time_ it was called. This comes from the default value of the third argument `cond` being set to `quote(._counter_ == 1)`.

As you might have guessed, `._counter_` is an internal variable that keeps track of how many times the method has been called. In this case, `ggtrace_inspect_return()` is giving us the return value from the method when it was first run, i.e., in the first group of the first panel. 

If we instead wanted to get the return value of `StatSmooth$compute_group` for the third group of the second panel, for example, we can do so in one of two ways:

1) Set the value of `cond` to an expression that evaluates to true for that panel and group:

```{r}
return_val_2_3_A <- ggtrace_inspect_return(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3)
)
```

2) Find the counter value when that condition is satisfied with `ggtrace_inspect_which()`, and then simply check for the value of `._counter_` back in `ggtrace_inspect_return()`:

```{r}
ggtrace_inspect_which(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3)
)

return_val_2_3_B <- ggtrace_inspect_return(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(._counter_ == 6)
)
```

These two approaches work the same:

```{r}
identical(return_val_2_3_A, return_val_2_3_B)
```

### **Capture**

Okay, so we know what `StatSmooth$compute_group` returns, but how does this return value change with different input? More generally put, how does `StatSmooth$compute_group` behave under different contexts?

We _could_ answer this by making a bunch of different plots using `geom_smooth()` and repeating the inspection workflow. Alternatively, we can capture a call to `StatSmooth$compute_group` and extract it as a function with `ggtrace_capture_fn()`:

```{r}
captured_fn_2_3 <- ggtrace_capture_fn(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3)
)
```

`captured_fn_2_3` is essentially a snapshot of the `compute_group` when it is called for the third group of the second panel. Simply calling `captured_fn_2_2` gives us the expected return value:

```{r}
identical(return_val_2_3_A, captured_fn_2_3())
```

But the true power of the "capture" workflow functions lies in the ability to interact with what has been captured. In the case of `ggtrace_capture_fn()`, the returned function has all of the arguments passed to it at its execution stored in the formals.

In other words, it is "pre-filled" with its original values, which we can inspect with `formals()`:

```{r}
formals(captured_fn_2_3)
```

This makes it very convenient for us to explore its behavior with different arguments passed to it.

For example, when `flipped_aes = TRUE`, we get `xmin` and `xmax` columns replacing `ymin` and `ymax`:

```{r}
head(captured_fn_2_3(flipped_aes = TRUE))
```

In this sense, we can effectively simulate what happens in `geom_smooth(orientation = "y")` without needing to construct an entirely different ggplot.

For another example, when we set the confidence interval to 10% with `level = 0.1`, the `ymin` and `ymax` values deviate less from the `y` value:

```{r}
head(captured_fn_2_3(level = 0.1))
```


Lastly, let's talk about the `data` variable we've been using inside the `cond` argument of some of these workflow functions. What is `data$group` and `data$PANEL`? How do you know what `data` looks like?

The answer is actually simple: it's an argument passed to `StatSmooth$compute_group`. We already saw it above but to make pull it out more explicitly, `data` looks like this for the third group of the second group:

```{r}
formals(captured_fn_2_3)$data
```

We see that `PANEL` and `group` columns conveniently give us information about the panel and group that the `compute_group` is doing calculations for.


### **Highjack**

Once we're satisfied about our understanding of how `StatSmooth$compute_group` works, we may want to test some hypotheses about what would happen if the `compute_group` method returned something else.

Let's revisit our examples from the Capture workflow. What if the third group of the second panel was calculating a more conservative confidence interval (`level = 0.1`)? What is this effect on the graphical output?

To answer this question, we use `ggtrace_highjack_return()` to have a method return an entirely different value, with help of `{rlang}`.

First we store the modified return value in some variable:

```{r}
modified_return_smooth <- captured_fn_2_3(level = 0.1)
```

Then we target the same group inside `cond` and pass the modified values to the `value` argument using `!!` and `rlang::expr()`

```{r}
ggtrace_highjack_return(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3),
  value = rlang::expr(!!modified_return_smooth)
)
```

The confidence band is nearly invisible for that fitted line!

Here's another example where we make the method fit predictions from a loess regression instead. This time, we in-line the calculation of the new return value:

```{r}
ggtrace_highjack_return(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3),
  value = rlang::expr(!!captured_fn_2_3(method = "loess"))
)
```

But that's not all! The `value` argument of `ggtrace_highjack_return()` exposes an internal function called `returnValue()` which simply returns the original return value. Computing on it allows on-the-fly modifications the graphical output.

For example, we can "intercept" the dataframe output, do data wrangling on it, and have the method return that instead. Here, we hack the data for the group to make it look like there's an absurd degree of heteroskedasticity:

```{r}
library(dplyr)
ggtrace_highjack_return(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3),
  value = quote({
    
    spread_seq <- seq(0, 10, length.out = nrow(returnValue()))
    returnValue() %>% 
      mutate(
        ymin = y - se * spread_seq[row_number()],
        ymax = y + se * spread_seq[row_number()]
      )
    
  })
)
```


## **Middle ground approach `with_ggtrace()`**

So far we've seen the low-level function `ggtrace()` and the high-level family of workflow functions `ggtrace_{action}_{value}()`. But once you get more familiar with ggplot internals more and start using `ggtrace()` to "hack into" the internals (moving from _learner_ to _developer_, in a sense), you might want both exploit both the power and convenience that `{ggtrace}` provides across these two designs.

For that we provide `with_ggtrace()`, which wraps around `ggtrace()` to give you access to its full power, while keeping its effects localized (e.g., no side-effects) and therefore more fitting for a functional programming workflow.

Like `ggtrace()`, you inject code into different steps of a method _as it runs_:

```{r}
with_ggtrace(
  x = p + facet_grid(year ~ drv),
  method = Layout$render,
  trace_steps = c(5, 8),
  trace_exprs = rlang::exprs(
    
    {
      # First, turn all panels except the 4th semi-transparent
      panels[-4] <- lapply(panels[-4], function(panel) {
        editGrob(panel, gp = gpar(alpha = .4))
      })
      # Second, give red outline and fill to 4th panel
      panels[[4]] <- gTree(children = gList(panels[[4]],
        rectGrob(x = 0.5, y = 0.5, width = 1, height = 1,
                 gp = gpar(col = "red", lwd = 5, fill = "red", alpha = 0.1))
      ))
    },
    
    {
      # Third, give it some emphasis by connecting to facet strips
      outline_rect <- rectGrob(
        x = 0.5, y = 0.5, width = 1, height = 1,
        gp = gpar(col = "red", lwd = 2, fill = NA)
      )
      plot_table <- gtable_add_grob(
        plot_table, outline_rect,
        t = 1, l = 2, b = 5, r = 2, z = Inf
      )
      plot_table <- gtable_add_grob(
        plot_table, outline_rect,
        t = 5, l = 2, b = 5, r = 7, z = Inf
      )
    }
    
  ),
  out = "g"
)
```

And like the workflow functions, you can have conditional traces that only evaluate when a condition is met, using `if` statements inside `trace_exprs` with `once = FALSE`.

```{r}
with_ggtrace(
  p,
  GeomRibbon$draw_group,
  trace_steps = -1,
  trace_exprs = quote({
    # Give gradient fill to the confidence bands for group 3
    if (data$group[1] == 3) {
      g_poly <- editGrob(
        g_poly,
        gp = gpar(fill = linearGradient(
          colours = c("purple", "skyblue", "pink"),
          stops = c(0, 0.5, 1),
          x1 = 0, x2 = 1, y1 = 0, y2 = 0
        ))
      )
    }
  }),
  once = FALSE,
  out = "g"
)
```

And of course, all of this is not limited to objects from the `{ggplot2}` package itself. You can have fun hacking extension packages as well!

```{r}
library(ggh4x)

# Example from - https://teunbrand.github.io/ggh4x/articles/Facets.html ----

## Base plot ====
p <- ggplot(mpg, aes(displ, hwy, colour = as.factor(cyl))) + geom_point() +
  labs(x = "Engine displacement", y = "Highway miles per gallon") +
  guides(colour = "none")

## Fixed-aspect plot with free & independent scales using `ggh4x::facet_grid2()` ====
p2 <- p +
  ggh4x::facet_grid2(vars(year), vars(drv), scales = "free", independent = "all") +
  theme_grey(base_size = 9) +
  theme(aspect.ratio = 1)

# Highjacking the plot's execution using `ggtrace::with_ggtrace()` ----
with_ggtrace(
  
  ## Argument 1: The ggplot to interact with durings its execution ====
  x = p2,
  
  ## Argument 2: The method to inject code into ====
  method = FacetGrid2$draw_panels,
  
  ## Argument 3: The step in the method right before `panels` and `axes` ====
  ## are merged into the `panel_table` <gtable> and shipped off
  trace_steps = 13, # See `ggbody(FacetGrid2$draw_panels)`
  
  ## Argument 4: The code injection to rotate the 2nd panel (1st row, 2nd column) ====
  trace_exprs = quote({
    
    ### FIRST, specify row/column of panel to target ####
    row <- 1
    col <- 2
    target_panel <- layout[layout$ROW == row & layout$COL == col, ]$PANEL
    left_axis    <- axes$left[[row, col]]
    bottom_axis  <- axes$bottom[[row, col]]
    
    ### SECOND, replace target panelwith a version of itself ####
    ### that has its axes attached to itself, using `grid::gTree()`
    panels[[target_panel]] <- gTree(
      
      ##### Argument `children`: a list (`grid::gTree()`) of three grobs:
      children = gList(
        
        ###### 1. The original panel itself
        panels[[target_panel]],
        
        ###### 2. Resized/reoriented left y-axis for that panel
        editGrob(
          left_axis,
          vp = viewport(
            x = unit(0.5, "npc") - grobWidth(left_axis) / 2,
            just = c("right")
          )
        ),
        
        ###### 3. Resized/repositioned bottom x-axis for that panel
        editGrob(
          bottom_axis,
          vp = viewport(
            y = unit(0.5, "npc") - grobHeight(bottom_axis) / 2,
            just = c("top")
          )
        )
        
      ),
      
      ##### Argument `vp`: Rotation for this combination of three grobs
      vp = viewport(width = .7, height = .7, angle = 45)
      
    )
    
    ### THIRD, "remove" the original axes for that panel ####
    axes$left[[row, col]]    <- zeroGrob() 
    axes$bottom[[row, col]]  <- zeroGrob()
    
  }),
  
  ## Argument `out`: The return value for `with_ggtrace()` ====
  ## "gtable" returns the graphical output after injecting the code
  out = "gtable" # you can also use the "g" shorthand
  
)
```
