---
title: "ggtrace"
output:
  md_document:
    toc: true
    toc_depth: 2
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
ragg_png = function(...) {
  ragg::agg_png(..., res = 300, units = "in")
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "ragg_png",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# **{ggtrace}**

#### **Programmatically explore, debug, and manipulate ggplot internals**

<!-- badges: start -->
[![](https://img.shields.io/badge/devel%20version-0.5.1-gogreen.svg)](https://github.com/yjunechoe/ggtrace)
<!-- badges: end -->

### **Installation**

You can install the development version from [GitHub](https://github.com/yjunechoe/ggtrace/) with:

``` r
# install.packages("remotes")
remotes::install_github("yjunechoe/ggtrace")
```

```{r setup}
library(ggtrace) # v0.5.1
```

More on the 📦 package website: [https://yjunechoe.github.io/ggtrace](https://yjunechoe.github.io/ggtrace)

## **Extending `base::trace()` with `ggtrace()`**

The low-level function `ggtrace()` is designed for interacting with functions and ggproto methods in the `{ggplot2}` ecosystem, from the "outside". 

Formally put, `ggtrace()` allows the user to inject arbitrary expressions (called **traces**) to functions and methods that are evaluated over the execution of a ggplot. When "triggered" by the evaluation of the ggplot, these traces may modify the resulting graphical output, or they may simply log their values to the "tracedump" for further inspection by the user. Check out the [FAQ vignette](https://yjunechoe.github.io/ggtrace/articles/FAQ.html) for more details.

Briefly, there are three key arguments to `ggtrace()`:

- `method`: what function/method to trace
- `trace_steps`: where in the body to inject expressions
- `trace_exprs` what expressions to inject

A simple example:

```{r}
dummy_fn <- function(x = 1, y = 2) {
  z <- x + y
  return(z)
}
dummy_fn()
```

Essentially, `ggtrace()` allows you to inject code into a function and safely/temporarily change its execution behavior. 

The following code injects the code `z <- z * 10` right as `dummy_fn` enters the third "step" in the body, _right before_ the line `return(z)` is ran.

```{r}
body(dummy_fn)[[3]]
```

```{r}
ggtrace(
  method = dummy_fn,
  trace_steps = 3L, # Before `return(z)` is ran
  trace_exprs = quote(z <- z * 10)
)
```

Note that the value of `trace_exprs` must be of type "language" (a quoted expression), the idea being that we are _injecting_ code to be evaluate inside the function when it is called. Often, providing the code wrapped in `quote()` suffices. For more complex injections see the [Expressions chapter of Advanced R](https://adv-r.hadley.nz/expressions.html)

After this `ggtrace()` call, the next time `dummy_fn` is called it is run with this injected code.

```{r}
# Returns 30 instead of 3
dummy_fn()
```

Essentially, `dummy_fn` ran with this following modified code just now:

```{r}
dummy_fn_traced <- function(x = 1, y = 2) {
  z <- x + y
  z <- z * 10 #< injected code!
  return(z)
}
dummy_fn_traced()
```

By default, traces created by `{ggtrace}` functions delete themselves after being triggered. You can also check whether a function is currently being traced with `is_traced()`.

```{r}
is_traced(dummy_fn)
```

`{ggtrace}` automatically logs the output of triggered trace to what we call **tracedumps**. For example, `last_ggtrace()` stores the output of the _last_ trace created by `ggtrace()`:

```{r}
# The value of `(z <- z * 10)` when it was ran
last_ggtrace() # Note that this is a list of length `trace_steps`
```

See the references section [**Extending base::trace()**](https://yjunechoe.github.io/ggtrace/reference/index.html#extending-base-trace-) for more functionalities offered by `ggtrace()`.

## **Workflows for interacting with ggplot internals**

```{r}
library(ggplot2) # v3.3.5
```

Admittedly, `ggtrace()` is a bit too clunky for interactive explorations of ggplot internals. To address this, we offer "workflow" functions in the form of `ggtrace_{action}_{value}()`. These are grouped into three workflows: Inspect, Capture, and Highjack.

**NOTE**: Making the most out of these workflow functions requires a hint of knowledge about ggplot internals, namely the fact that **ggproto objects** like [Stat](https://ggplot2.tidyverse.org/reference/ggplot2-ggproto.html#stats) and [Geom](https://ggplot2.tidyverse.org/reference/ggplot2-ggproto.html#geoms) exists, and that these [ggprotos have methods](https://ggplot2-book.org/spring1.html#methods) that step in at different parts of the ggplot build/render pipeline to modify the data. If you are completely new to these concepts, you should at least watch [Thomas Lin Pedersen](https://twitter.com/thomasp85)'s talk on [Extending your ability to extend ggplot2](https://www.rstudio.com/resources/rstudioconf-2020/extending-your-ability-to-extend-ggplot2/) before proceeding.

### **Walkthrough with `geom_smooth()`**

Say we want to learn more about how `geom_smooth()` layer works, exactly

```{r}
geom_smooth()
```

To do this, we're going to adopt the example from the [ggplot2 internals chapter of the ggplot book](https://ggplot2-book.org/internals.html)

```{r}
p <- ggplot(mpg, aes(displ, hwy, color = drv)) + 
  geom_point(position = position_jitter(seed = 1116)) +
  geom_smooth(method = "lm", formula = y ~ x) + 
  facet_wrap(vars(year)) + 
  ggtitle("A plot for expository purposes")
p
```

Let's focus on the Stat ggproto. We see that `geom_smooth()` uses the `StatSmooth` ggproto

```{r}
class( geom_smooth()$stat )

identical(StatSmooth, geom_smooth()$stat)
```

The bulk of the work by a Stat is done in the `compute_*` family of methods, which are essentially just functions. We'll focus on `compute_group` here:

```{r}
# ggproto methods wrap over the actual function and print extra info
class( StatSmooth$compute_group )

# Use `get_method` to pull out just the function component
class( get_method(StatSmooth$compute_group) )

# StatSmooth inherits `compute_layer`/`compute_panel` and defines `compute_group`
get_method_inheritance(StatSmooth)
```

### **Inspect**

Here we introduce our first workflow function `ggtrace_inspect_n()`, which takes a ggplot as the first argument and a ggproto method as the second argument, returning the number of times the ggproto method has been called in the ggplot's evaluation:

```{r}
ggtrace_inspect_n(x = p, method = StatSmooth$compute_group)
```

As we might have guessed, `StatSmooth$compute_group` is called for each fitted line (each group) in the plot. But if `StatSmooth$compute_group` is essentially a function, what does it return?

We can answer that with another workflow function `ggtrace_inspect_return()`, which shares a similar syntax:

```{r}
return_val <- ggtrace_inspect_return(x = p, method = StatSmooth$compute_group)

dim(return_val)

head(return_val)
```

Note that `ggtrace_inspect_return()` only gave us 1 dataframe, corresponding to the return value of `StatSmooth$compute_group` the _first time_ it was called. This comes from the default value of the third argument `cond` being set to `quote(._counter_ == 1)`.

Here, `._counter_` is an internal variable that keeps track of how many times the method has been called. It's available for all workflow functions and you can read more in the [**Tracing context**](https://yjunechoe.github.io/ggtrace/reference/ggtrace_inspect_return.html#tracing-context) section of the docs.

If we instead wanted to get the return value of `StatSmooth$compute_group` for the third group of the second panel, for example, we can do so in one of two ways:

  1. Set the value of `cond` to an expression that evaluates to true for that panel and group:
  
      ```{r}
      return_val_2_3_A <- ggtrace_inspect_return(
        x = p,
        method = StatSmooth$compute_group,
        cond = quote(data$PANEL[1] == 2 && data$group[1] == 3)
      )
      ```

  2. Find the counter value when that condition is satisfied with `ggtrace_inspect_which()`, and then simply check for the value of `._counter_` back in `ggtrace_inspect_return()`:
  
      ```{r}
      ggtrace_inspect_which(
        x = p,
        method = StatSmooth$compute_group,
        cond = quote(data$PANEL[1] == 2 && data$group[1] == 3)
      )
      
      return_val_2_3_B <- ggtrace_inspect_return(
        x = p,
        method = StatSmooth$compute_group,
        cond = 6L # shorthand for `quote(._counter_ == 6L)`
      )
      ```

These two approaches work the same:

```{r}
identical(return_val_2_3_A, return_val_2_3_B)
```

### **Capture**

Okay, so we know what `StatSmooth$compute_group` returns, but how does this return value change with different input? More generally put, how does `StatSmooth$compute_group` behave under different contexts?

We _could_ answer this by making a bunch of different plots using `geom_smooth()` and repeating the inspection workflow. Alternatively, we can capture a call to `StatSmooth$compute_group` and extract it as a function with `ggtrace_capture_fn()`:

```{r}
captured_fn_2_3 <- ggtrace_capture_fn(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3)
)
```

`captured_fn_2_3` is essentially a snapshot of the `compute_group` when it is called for the third group of the second panel. Simply calling `captured_fn_2_3` gives us the expected return value:

```{r}
identical(return_val_2_3_A, captured_fn_2_3())
```

But the true power of the "capture" workflow functions lies in the ability to interact with what has been captured. In the case of `ggtrace_capture_fn()`, the returned function has all of the arguments passed to it at its execution stored in the formals.

In other words, it is "pre-filled" with its original values, which we can inspect with `formals()`:

```{r, R.options = list(width = 70)}
# Just showing their type/class for space
sapply( formals(captured_fn_2_3) , class)
```

This makes it very convenient for us to explore its behavior with different arguments passed to it.

For example, when `flipped_aes = TRUE`, we get `xmin` and `xmax` columns replacing `ymin` and `ymax`:

```{r}
head( captured_fn_2_3(flipped_aes = TRUE) )
```

In this sense, we can effectively simulate what happens in `geom_smooth(orientation = "y")` without needing to construct an entirely different ggplot.

For another example, when we set the confidence interval to 10% with `level = 0.1`, the `ymin` and `ymax` values deviate less from the `y` value:

```{r}
head( captured_fn_2_3(level = 0.1) )
```

Lastly, let's talk about the `data` variable we've been using inside the `cond` argument of some of these workflow functions. What is `data$group` and `data$PANEL`? How do you know what `data` looks like?

The answer is actually simple: it's an argument passed to `StatSmooth$compute_group`. We saw earlier that it's stored in `formals(captured_fn_2_3)`, but to target it explicitly we can also use `ggtrace_inspect_args()`:

```{r}
args_2_3 <- ggtrace_inspect_args(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3)
)

identical(names(args_2_3), names(formals(captured_fn_2_3)))

args_2_3$data
```

We see that `PANEL` and `group` columns conveniently give us information about the panel and group that `compute_group` is doing calculations over.


### **Highjack**

Once we have some understanding of how `StatSmooth$compute_group` works, we may want to test some hypotheses about what would happen to the resulting graphical output if the method returned something else.

Let's revisit our examples from the Capture workflow. What if the third group of the second panel calculated a more conservative confidence interval (`level = 0.1`)? What is this effect on the graphical output?

To answer this question, we use `ggtrace_highjack_return()` to have a method return an entirely different value by supplying it an [expression](https://adv-r.hadley.nz/expressions.html).

First we store the modified return value in some variable:

```{r}
modified_return_smooth <- captured_fn_2_3(level = 0.1)
```

Then we target the same group inside `cond` and pass `modified_return_smooth` to the `value` argument _as an expression_, using `substitute()`:

```{r}
ggtrace_highjack_return(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3),
  value = substitute(modified_return_smooth)
          # or `rlang::expr(!!modified_return_smooth)`
)
```

The confidence band is now nearly invisible for that fitted line because it's only capturing a 10% confidence interval!

Here's another example where we make the method fit predictions from a loess regression instead. To achieve this directly, we use `ggtrace_highjack_args()` here and set the `values` to `list(method = "loess")`:

```{r}
ggtrace_highjack_args(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3),
  values = list(method = "loess")
)
```

Lastly, `ggtrace_highjack_return()` exposes an internal function called `returnValue()` in the `value` argument, which simply returns the original return value. Computing on it allows on-the-fly modifications to the graphical output.

For example, we can "intercept" the dataframe output, do data wrangling on it, and have the method return that instead. Here, we hack the data for the group to make it look like there's an absurd degree of heteroskedasticity:

```{r, message=FALSE, warning=FALSE}
library(dplyr) # v1.0.8

ggtrace_highjack_return(
  x = p,
  method = StatSmooth$compute_group,
  cond = quote(data$PANEL[1] == 2 && data$group[1] == 3),
  value = quote({
    
    returnValue() %>% 
      mutate(
        ymin = y - se * seq(0, 10, length.out = n()),
        ymax = y + se * seq(0, 10, length.out = n())
      )
    
  })
)
```


## **Middle-ground approach `with_ggtrace()`**

So far we've seen the low-level function `ggtrace()` and the high-level family of workflow functions `ggtrace_{action}_{value}()`. But once you get more familiar with ggplot internals and start using `ggtrace()` to "hack into" the internals (moving from _learner_ to _developer_, in a sense), you might want both exploit both the power and convenience that `{ggtrace}` provides across these two designs.

For that we provide `with_ggtrace()`, which wraps around `ggtrace()` to give you access to its full power, while keeping its effects localized (e.g., no side-effects) and therefore making it more fitting for a functional programming workflow.

Like `ggtrace()`, you can inject code into different steps of a method _as it runs_:

```{r}
with_ggtrace(
  x = p + facet_grid(year ~ drv),
  method = Layout$render,
  trace_steps = c(5L, 8L),
  trace_exprs = rlang::exprs(
    
    {
      # First, turn all panels except the 4th semi-transparent
      panels[-4] <- lapply(panels[-4], function(panel) {
        editGrob(panel, gp = gpar(alpha = .4))
      })
      # Second, give red outline and fill to 4th panel
      panels[[4]] <- gTree(children = gList(panels[[4]],
        rectGrob(x = 0.5, y = 0.5, width = 1, height = 1,
                 gp = gpar(col = "red", lwd = 5, fill = "red", alpha = 0.1))
      ))
    },
    
    {
      # Third, give it some emphasis by connecting to facet strips
      outline_rect <- rectGrob(
        x = 0.5, y = 0.5, width = 1, height = 1,
        gp = gpar(col = "red", lwd = 2, fill = NA)
      )
      plot_table <- gtable_add_grob(
        plot_table, outline_rect,
        t = 1, l = 2, b = 5, r = 2, z = Inf
      )
      plot_table <- gtable_add_grob(
        plot_table, outline_rect,
        t = 5, l = 2, b = 5, r = 7, z = Inf
      )
    }
    
  ),
  out = "g"
)
```

And like the workflow functions, you can have conditional traces that only evaluate when a condition is met, using `if` statements inside `trace_exprs` with `once = FALSE`.

```{r}
with_ggtrace(
  p,
  GeomRibbon$draw_group,
  trace_steps = -1L,
  trace_exprs = quote({
    # Give gradient fill to the confidence bands for third group
    if (data$group[1] == 3) {
      g_poly <- editGrob(
        g_poly,
        gp = gpar(fill = linearGradient(
          colours = c("purple", "skyblue", "pink"),
          stops = c(0, 0.5, 1),
          x1 = 0, x2 = 1, y1 = 0, y2 = 0
        ))
      )
    }
  }),
  once = FALSE, # to test `if` for all calls to method
  out = "g"
)
```

And of course, all of this is not limited to objects from the `{ggplot2}` package itself. You can have fun hacking extension packages as well!

```{r}
library(ggh4x) # v0.2.1

# Example from - https://teunbrand.github.io/ggh4x/articles/Facets.html ----

## Base plot ====
p <- ggplot(mpg, aes(displ, hwy, colour = as.factor(cyl))) + geom_point() +
  labs(x = "Engine displacement", y = "Highway miles per gallon") +
  guides(colour = "none")

## Fixed-aspect plot w/ independent scales using `ggh4x::facet_grid2()` ====
p2 <- p +
  ggh4x::facet_grid2(vars(year), vars(drv), scales = "free", independent = "all") +
  theme_grey(base_size = 9) +
  theme(aspect.ratio = 1)

# Highjacking the plot's execution using `ggtrace::with_ggtrace()` ----
with_ggtrace(
  
  ## Argument 1: The ggplot to interact with durings its execution ====
  x = p2,
  
  ## Argument 2: The method to inject code into ====
  method = ggh4x::FacetGrid2$draw_panels,
  
  ## Argument 3: The step in the method right before `panels` and `axes` ====
  ## are merged into the `panel_table` <gtable> and shipped off
  trace_steps = 13, # See `ggbody(FacetGrid2$draw_panels)`
  
  ## Argument 4: The code injection to rotate the 2nd panel ====
  trace_exprs = quote({
    
    ### FIRST, specify row/column of panel to target ####
    row <- 1
    col <- 2
    target_panel <- layout[layout$ROW == row & layout$COL == col, ]$PANEL
    left_axis    <- axes$left[[row, col]]
    bottom_axis  <- axes$bottom[[row, col]]
    
    ### SECOND, replace target panel with a version of itself ####
    ### that has axes attached to it, using `grid::gTree()`
    panels[[target_panel]] <- gTree(
      
      ##### Argument `children`: a list (`grid::gList()`) of three grobs:
      children = gList(
        
        ###### 1. The original panel itself
        panels[[target_panel]],
        
        ###### 2. Resized/repositioned left y-axis for that panel
        editGrob(
          left_axis,
          vp = viewport(
            x = unit(0.5, "npc") - grobWidth(left_axis) / 2,
            just = c("right")
          )
        ),
        
        ###### 3. Resized/repositioned bottom x-axis for that panel
        editGrob(
          bottom_axis,
          vp = viewport(
            y = unit(0.5, "npc") - grobHeight(bottom_axis) / 2,
            just = c("top")
          )
        )
        
      ),
      
      ##### Argument `vp`: Rotation for this combination of three grobs
      vp = viewport(width = .7, height = .7, angle = 45)
      
    )
    
    ### THIRD, "remove" the original axes for that panel ####
    axes$left[[row, col]]    <- zeroGrob() 
    axes$bottom[[row, col]]  <- zeroGrob()
    
  }),
  
  ## Argument `out`: The return value for `with_ggtrace()` ====
  ## "gtable" returns the graphical output after injecting the code
  out = "gtable" # you can also use the "g" shorthand
  
)
```

## **Case study: Data-driven legends**

Suppose we want to change the fill legend key from squares to the actual violins drawn for each category:

```{r}
violin_plot <- dplyr::starwars %>% 
  filter(species == "Human", !is.na(height)) %>% 
  ggplot(aes(gender, height)) +
  geom_violin(aes(fill = gender)) +
  theme(
    aspect.ratio = 1,
    legend.key.size = unit(1, "in"),
    legend.key = element_rect(color = "grey", fill = NA)
  )

violin_plot
```

We know that each violin is drawn by `GeomViolin$draw_group`, called for each group in the plot.

```{r}
ggtrace_inspect_n(violin_plot, GeomViolin$draw_group)
```

Let's grab the plotted violins with `ggtrace_inspect_return()`

```{r}
violins <- lapply(1:2, ggtrace_inspect_return,
                  x = violin_plot, method = GeomViolin$draw_group)
violins
```

Then, we can use `ggtrace_highjack_return()` to highjack the return value of the key drawing method `GeomViolin$draw_key` such that it returns the output from the `draw_group` method instead. We set `cond = 1:2` and `value = substitute(violins[[._counter_]])` so that the `draw_key` method returns the first violin as the first legend key, and the second violin as the second legend key.

```{r}
ggtrace_highjack_return(
  violin_plot,
  GeomViolin$draw_key,
  cond = 1:2, # or `cond = TRUE` since method is called twice total
  value = substitute(violins[[._counter_]])
)
```

Note that violins in the legend are drawn with respect to the panel coordinate system, because that's how the `draw_group` method does things. For a cleaner look, we use the defined function `center_and_resize()` which modifies the violin grob using `grid::editGrob()` before passing them off to `ggtrace_highjack_return()`.

```{r}
library(grid)
center_and_resize <- function(grob) {
  x_range <- range(unclass(grob$x))
  y_range <- range(unclass(grob$y))
  width <- diff(x_range)
  height <- diff(y_range)
  resizing <- (1 - 0.2)/max(width, height)
  editGrob(grob, vp = viewport(
    xscale = x_range, yscale = y_range,
    width = width * resizing, height = height * resizing
  ))
}
violins_centered_resized <- lapply(violins, center_and_resize)

violin_plot_highjacked <- ggtrace_highjack_return(
  violin_plot,
  GeomViolin$draw_key,
  cond = 1:2,
  value = substitute(violins_centered_resized[[._counter_]])
)
violin_plot_highjacked
```

To keep this effect while keeping the violin_plot a ggplot object, you can create a custom key glyph function that inspects the fill color of the original legend keys and returns the corresponding violin.

```{r}
names(violins_centered_resized) <- c("#F8766D", "#00BFC4")

violin_plot_keyglyph <- dplyr::starwars %>% 
  filter(species == "Human", !is.na(height)) %>% 
  ggplot(aes(gender, height)) +
  geom_violin(
    aes(fill = gender),
    key_glyph = function(data, params, size) {   #<
      violins_centered_resized[[data$fill]]      #<  Custom `key_glyph` function
    }                                            #<
  ) +
  theme(
    aspect.ratio = 1,
    legend.key.size = unit(1, "in"),
    legend.key = element_rect(color = "grey", fill = NA)
  )
```

Note the difference in output between the two methods. While highjack workflow functions offer a convenient way of modifying a ggplot, it forces the conversion of a ggplot object into a grob.

```{r, error = TRUE}
class(violin_plot_highjacked)
violin_plot_highjacked +
  labs(title = "Violin plot with data-driven legend key glyphs")

class(violin_plot_keyglyph)
violin_plot_keyglyph +
  labs(title = "Violin plot with data-driven legend key glyphs")
```

The lesson here is that the highjack workflow (and the {ggtrace} package, more broadly) is the means, not the end. Users are encouraged to generalize solutions beyond one-off hacks using the toolkit of workflow functions to determine the appropriate point of extension.
