---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# **{ggtrace}**

<!-- badges: start -->
<!-- badges: end -->

An experimental package for programmatically debugging ggproto methods with `trace()`.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yjunechoe/ggtrace")
```

## Usage

```{r setup}
library(ggtrace)
library(ggplot2) # v3.3.5
```

## Example 1 - `compute_layer` method from `PositionJitter` {ggplot2}

### Step 1. Make plot

```{r ex-1-setup}
jitter_plot <- ggplot(diamonds[1:1000,], aes(cut, depth)) +
  geom_point(position = position_jitter(width = 0.2, seed = 2021))
jitter_plot
```

### Step 2. Inspect callstack of the ggproto of interest

```{r ex-1-ggbody}
ggbody(PositionJitter$compute_layer)
```

### Step 3a. `ggtrace()` - with expressions wrapped in `head()`

```{r ex-1a-ggtrace, fig.show='hide'}
ggtrace(
  method = PositionJitter$compute_layer,
  trace_steps = c(1, 1, 9, 12),
  trace_exprs = rlang::exprs(
    head(data),        # What does the data passed in look like?
    params,            # What do the initial parameters look like?
    head(dummy_data),  # What is `dummy_data` defined at Step 8?
    head(~line)        # What does the last line evaluate to?
                       # - i.e., what is returned by the method?
  )
)

# plot not printed to save space
jitter_plot
```

### Step 4a. Inspect trace dumnp

```{r ex-1a-last}
last_ggtrace()
```

### Step 3b. `ggtrace()` - with `.print = FALSE` (recommended)

```{r ex-1b-ggtrace, fig.show='hide'}
ggtrace(
  method = PositionJitter$compute_layer,
  trace_steps = c(1, 1, 12),
  trace_exprs = rlang::exprs(
    data,
    dummy_data,
    ~line
  ),
  .print = FALSE
)

# plot not printed to save space
jitter_plot
```

### Step 4b. Inspect trace dump

```{r ex-1b-last}
jitter_tracedump <- last_ggtrace()
lapply(jitter_tracedump, nrow)
lapply(jitter_tracedump, head)
```

## Example 2 - `draw_group` method from `GeomSmooth` {ggplot2}

### Step 1. Make plot

```{r ex-2-setup}
smooth_plot <- ggplot(mtcars, aes(mpg, hp)) +
  geom_point() +
  stat_smooth(method = "lm")
smooth_plot
```

### Step 2. Inspect callstack of the ggproto of interest

```{r ex-2-ggbody}
ggbody(GeomSmooth$draw_group)
```

### Step 3. `ggtrace()` - get gList

```{r ex-2-ggtrace, fig.show='hide'}
ggtrace(
  method = GeomSmooth$draw_group,
  trace_steps = 7,
  trace_exprs = quote(~line),
  .print = FALSE
)

# plot not printed to save space
smooth_plot
```

### Step 4. Inspect trace dump

```{r ex-2-last}
smooth_tracedump <- last_ggtrace()
smooth_tracedump
str(smooth_tracedump[[1]])
```


## Example 3 - `compute_group` method from `StatSina` {ggforce}

### Step 1. Make plot

```{r ex-3-setup}
library(ggforce)

sina_plot <- ggplot(diamonds[diamonds$cut == "Ideal",][1:50,], aes(cut, depth)) +
  geom_violin() +
  geom_sina(seed = 2021)
sina_plot
```

### Step 2. Inspect callstack of the ggproto of interest

```{r ex-3-ggbody}
ggbody(StatSina$compute_group)
```

### Step 3. `ggtrace()` - with one expression evaluated at multiple steps

```{r ex-3-ggtrace, fig.show='hide'}
ggtrace(
  method = StatSina$compute_group,
  trace_steps = c(1, 8),
  trace_exprs = quote(data),
  .print = FALSE
)

# plot not printed to save space
sina_plot
```

### Step 4. Inspect trace dump

```{r ex-3-last}
sina_tracedump <- last_ggtrace()
waldo::compare(sina_tracedump[[1]], sina_tracedump[[2]])
```
